---
title: "Density Dependent Population Regulation"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Density Dependent Population Regulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r doc_setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(landsim)
set.seed(42)
```

This document will go through
several ways of implementing density-dependent population regulation.

Here's the habitat we'll work with.
Note that units are in meters, and the resolution of the raster is 100m.
```{r setup_layer, fig.cap="The (random) habitat.", fig.height=4, fig.width=4}
pop <- make_population(
            habitat = random_habitat(),
            inaccessible.value = NA,
            uninhabitable.value = NA,
            genotypes = c("aa","aA","AA"),
            N = 0
        )
pop$N[,"aa"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/4)
pop$N[,"aA"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/2)
pop$N[,"AA"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/4)
plot(pop$habitat)
```

Here's the basic, default demography:
```{r basic_demog}
basic.migr <- migration(
                    kern = "gaussian",
                    sigma = 300,
                    radius = 1000,
                    normalize = 1
             )
basic.demog <- demography(
        prob.seed = 0.05,
        fecundity = 200,
        prob.germination = 0.4,
        prob.survival = 0.6,
        pollen.migration = basic.migr,
        seed.migration = basic.migr,
        genotypes = c("aa","aA","AA")
    )
```

Of course, 
this demography has no dependence on the current state of the population,
so the population numbers will grow without bound:
without taking into account migration,
the mean number of offspring is 
$r basic.demog[["prob.seed"]]` × `r basic.demog[["fecundity"]]` × `r basic.demog[["prob.germination"]]` + `r basic.demog[["prob.survival"]]` 
= `r basic.demog[["prob.seed"]] * basic.demog[["fecundity"]] * basic.demog[["prob.germination"]] + basic.demog[["prob.survival"]]`$
```{r show_unconstrained_growth}
demog <- setup_demography( basic.demog, pop )
sim <- simulate_pop( pop, demog, times=seq(0,10,length.out=11),
                summaries=list( totals=function(N){colSums(N)} )
            )
matplot(sim$summaries[["totals"]],type='l',lty=1, log='y',ylab='number of individuals')
legend("bottomright",lty=1,col=1:3,legend=pop$genotypes)
```


# Resource competition: available space

First, we could make the probability of germination (i.e., establishment)
depend on the nearby density of individuals.
Below, we set the probability of germination to $p=p_0/(1+n/K)$, 
where $n$ is a local average number of individuals per raster cell;
this is stable if $`r basic.demog[["prob.seed"]]` \times `r basic.demog[["fecundity"]]` \times p + `r basic.demog[["prob.survival"]]` = 1$, 
which happens if if $p=`r round((1-basic.demog[["prob.survival"]])/(basic.demog[["prob.seed"]]*basic.demog[["fecundity"]]),2)`$,
or, with $p_0 = `r basic.demog[["prob.germination"]]`$,
if $n=`r round(basic.demog[["prob.germination"]]/((1-basic.demog[["prob.survival"]])/(basic.demog[["prob.seed"]]*basic.demog[["fecundity"]]))-1,2)` K$.
```{r prob_germ}
demog <- basic.demog
demog$prob.germination <- vital(
                    function (N,...) {
                        out <- r0 / ( 1 + migrate(competition,x=rowSums(N))/K )
                        cbind( aa=out, aA=out, AA=out )
                    },
                    r0 = 0.4,
                    K = values(pop$habitat)[pop$habitable]/5,
                    competition = migration(
                                kern="gaussian",
                                sigma=200,
                                radius=400,
                                normalize=1
                        )
                )
```

This puts a contraint on the population:
```{r plot_prob_germ}
demog <- setup_demography( demog, pop )
sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=101),
                summaries=list( totals=function(N){colSums(N)} )
            )
matplot(sim$summaries[["totals"]],type='l',lty=1, log='y', ylab="number of individuals")
legend("bottomright",lty=1,col=1:3,legend=pop$genotypes)
```
```{r anim_prob_germ, fig.show="animate", aniopts="controls", interval=0.2}
plot(sim,pop,pause=FALSE)
```


# Density-dependent death rate

Now, we'll try a different tack, making death rates depend on local densities.
Since death occurs after reproduction,
we need the mean number of offspring to be less than one
for this to work.
```{r prob_death}
demog <- basic.demog
demog$prob.seed <- 0.01
demog$prob.survival <- vital(
                    function (N,...) {
                        out <- s0 / ( 1 + migrate(competition,x=rowSums(N))/K )
                        cbind( aa=out, aA=out, AA=out )
                    },
                    s0 = 0.6,
                    K = values(pop$habitat)[pop$habitable]/3,
                    competition = migration(
                                kern="gaussian",
                                sigma=200,
                                radius=400,
                                normalize=1
                        )
                )
```

Here's what that looks like:
```{r plot_prob_death}
demog <- setup_demography( demog, pop )
sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=101),
                summaries=list( totals=function(N){colSums(N)} ) )
matplot(sim$summaries[["totals"]],type='l',lty=1, log='y')
legend("bottomright",lty=1,col=1:3,legend=pop$genotypes)
```
```{r anim_prob_death, fig.show="animate", aniopts="controls", interval=0.2}
plot(sim,pop,pause=FALSE)
```
