---
title: "Simulations on large rasters."
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r doc_setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
devtools::load_all("..")
set.seed(42)
this.plot <- function (x,nr=1,zlim=c(0,max(values(x),na.rm=TRUE)),...) {
    plot(x,nr=nr,zlim=zlim,...)
}
```

**How well do the simulation methods work with larger rasters?**

Here are the demographic parameters:
```{r fancy_setup}
germination.args <- list( 
             r0 = 0.01,
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1
                                 )
         )
germination_fun <- function (N) {
    germination.args$r0 / ( 1 + migrate(N,germination.args$competition)/carrying.capacity )
}

this.demography <- demography(
        prob.seed = 0.2,
        fecundity = 200,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 100,
                            radius = 1000,
                            normalize = NULL
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 20,
                            radius = 400,
                            normalize = 1
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") ),
        germination.ags = germination.args,
        carrying.capacity=habitat
    )
```

Raster sizes
============

We will time a generation on rasters of different sizes.
Here is a function to create a $k$-kilometer square random raster.
(But note that the base units of the raster are in meters.)
```{r habitat_setup}
make_habitat <- function (k) {
    habitat <- raster(xmn=-k*1000/2, xmx=k*1000/2, ymn=-k*1000/2, ymx=k*1000/2, 
          resolution=100,
          crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
    values(habitat) <- sample( 100*c(1,2,5,NA), length(habitat), replace=TRUE )
    migrate(habitat,kern="gaussian",sigma=200,radius=1000)
}
```

Here's timing for one generation on rasters of various sizes:
```{r init_new_pop}
sizes <- 4^(0:3)
results <- lapply( sizes, function (k) {
                  habitat <- make_habitat(k)
                  N <- do.call( brick, list(habitat)[rep(1,length(this.demography$genotypes))] )
                  names(N) <- genotypes
                  gen.time <- system.time( NN <- generation(N,this.demography) )
                  return( list( time=gen.time, N=N, NN=NN ) )
          } )
```

Now, looking at the results:
```{r show_timing_results}
rbind( size=sizes, sapply( results, "[[", "time" ) )
```
