---
title: "Long-tailed migration"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(landsim)
# helper plotting function
pl <- function (z,zlim=range(0,z,finite=TRUE),...) {
    hab <- do.call(stack,list(habitat)[rep(1,NCOL(z))]); 
    values(hab)[habitable] <- z; 
    plot(hab,nr=1,zlim=zlim,legend.width=2,legend.mar=12,...) 
}
```

We'd like to run long-distance migration on a disconnected habitat 
where much of the range is inaccessible,
like this one:
```{r setup_habitat}
set.seed(42)
habitat <- raster::raster(xmn=-diam, xmx=diam, ymn=-diam, ymx=diam, 
      resolution=res,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
raster::values(habitat) <- pmax(-10,pmin(20,rcauchy(raster::ncell(habitat))))
habitat <- 20*migrate_raster( habitat, kern=kern, sigma=sigma, radius=radius )
values(habitat)[values(habitat) < (-5)] <- NA
plot(habitat)
```
We'll use negative values as accessible but inhospitable locations,
and NA values as inaccessible locations.

The problem is that the number of nonzero entries in the migration matrix
grows quadratically with the radius of the migration kernel.
To deal with this, we can use a smaller-radius migration matrix,
applied many times.
To do this we will

1. temporarily keep track of migration to *accessible* but not *habitable* locations
2. assign weights to different numbers of applications of the kernel

In general, the migration operator we want is
\[
    x \mapsto \sum_{n=0}^m w_n M^n ,
\]
with where $\sum w_n = 1$, and $M$ is some kind of fairly short-distance kernel.
Here's the setup:
```{r setup_migration}
accessible <- ( !is.na(values(habitat)) )
habitable <- ( accessible & values(habitat) > 0 )
# positions of habitable cells in vector of accessible cells
habitable.inds <- cumsum(accessible)[habitable]
M <- migration_matrix( habitat, kern="gaussian", sigma=100, radius=400,
                      from=which(accessible), to=which(accessible) )
x <- rpois( sum(habitable), values(habitat)[habitable] )
weighted_migrate <- function (n.weights) {
    Mnx <- numeric(sum(accessible))
    Mnx[habitable.inds] <- x
    out <- n.weights[1]*Mnx[habitable.inds]
    for (n in seq_along(n.weights)[-1]) {
        Mnx <- M %*% Mnx
        out <- out + n.weights[n]*Mnx[habitable.inds]
    }
    return(out)
}
```

With $w_n = (0,0,\ldots,1)$, this should look just like a Gaussian smooth,
with standard deviation equal to $\sigma \sqrt{m}$:
```{r gaussian}
n.weights <- c(rep(0,10),1)
pl( cbind( x=x, Mx=weighted_migrate(n.weights) ) )
```

If we take the smoothing operator as a geometric sum,
\[
    x \mapsto \frac{p}{1-(1-p)^{n=1}} \sum_{n=0}^m (1-p)^n M^n ,
\]
with $p$ fairly small, 
then we get a long-tailed distribution:

```{r cauchy}
p <- 1/10
n.weights <- (1-p)^(0:50)
n.weights <- n.weights/sum(n.weights)
pl( cbind( x=x, Mx=weighted_migrate(n.weights) ) )
```

