---
title: "Migration methods"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(testthat)
library(landsim)
```

**Note:** 
if `x` is a raster, `as.matrix(x)` is equal to the *transpose* of `matrix(values(x),nrow=nrow(x))`.
We want to access `values` directly, so we'll stay away from `as.matrix()`.


We'll start with a very simple example raster:
```{r simple_setup}
set.seed(42)
habitat <- raster(xmn=0, xmx=5, ymn=0, ymx=5, 
      resolution=1,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(habitat) <- sample( c(1,2), length(habitat), replace=TRUE )
habitat.NA <- habitat
values(habitat.NA)[c(1,12)] <- NA
layout(t(1:2))
plot(habitat,main="habitat",zlim=c(0,2))
plot(habitat.NA,main="habitat, with missings",zlim=c(0,2))
```

And, we'll need this:
```{r more_setup}
na_to_zero <- function (x) {
    x[is.na(x)] <- 0
    return(x)
}
```

Migration is just smoothing.
Let's make sure we understand what it does with boundaries,
and with missing values.

We'll use a weighting matrix that assigns weight $1/2$ to the center and weight $1/16$
to the surrounding 8 cells.
```{r weight_matrix}
w <- matrix(1/16,nrow=3,ncol=3)
w[2,2] <- 1/2
w
```

Types of boundary
=================

We would like to have two types of boundary:
**absorbing** boundaries where migrants try to go, but are killed,
and **non-boundaries** where migrants just don't try to go.
(The latter is a particular sort of reflecting boundary)
We may also want to treat the **internal** boundaries (`NA` cells)
and the **external** boundaries (border of the raster)
differently.

We will often want to zero out the `NA`s before computation,
but need to keep them around in some form to distinguish temporarily empty cells
from inhospitable ones.


Summary and usage
-----------------

`migration_matrix()`:
When precomputing a migration matrix,
set absorbing boundary elements to `accessible`,
but do not include them in `from` (and `to`).
Mark reflecting (non-)boundaries as not `accessible`.
External boundaries will be reflecting,
so if they should be absorbing, you need to extend the raster and set values appropriately
(see examples below).

`migrate_raster()`:
Currently, migration with `migrate_raster()`, which uses `raster::focal()`,
only implements absorbing boundaries.
Reflecting boundaries using raster-based methods requires using the much-slower `weighted.sum()` function;
see below for how to do this.


**Guts:** Absorbing internal and external boundaries
------------------------------------------

First, suppose we want migrants to exit at all boundaries.
`focal()` allows this by padding the matrix with zeros,
and using `na.rm=TRUE`.
```{r simple_focal}
hf1 <- focal( habitat, w=w, pad=TRUE, na.rm=TRUE, padValue=0 )
hf1.NA <- focal( habitat.NA, w=w, pad=TRUE, na.rm=TRUE, padValue=0 )
layout(t(1:2))
plot(hf1,main="habitat",zlim=c(0,2))
plot(hf1.NA,main="habitat, with missings",zlim=c(0,2))
```

Here's the same computation:
```{r check_simple}
hmat <- matrix(values(habitat),nrow=nrow(habitat),ncol=ncol(habitat))
hmat.NA <- matrix(values(habitat.NA),nrow=nrow(habitat),ncol=ncol(habitat))
expect_equivalent(as.numeric(hmat),values(habitat))
expect_equivalent(as.numeric(hmat.NA),values(habitat.NA))

hmat.pad <- rbind( 0, cbind( 0, hmat, 0 ), 0 )
hmat.NA.pad <- rbind( 0, cbind( 0, hmat.NA, 0 ), 0 )
hmf.NA <- hmf <- 0*hmat
for (dx in c(-1,0,1)) { 
    for (dy in c(-1,0,1)) {
        hmf <- hmf + hmat.pad[dx+1+(1:nrow(hmf)),dy+1+(1:ncol(hmf))]*w[2+dx,2+dy]
        add.these <- hmat.NA.pad[dx+1+(1:nrow(hmf)),dy+1+(1:ncol(hmf))]
        hmf.NA <- hmf.NA + ifelse(is.na(add.these),0,add.these)*w[2+dx,2+dy]
    }
}

expect_equivalent(as.numeric(hmf),values(hf1))
expect_equivalent(as.numeric(hmf.NA),values(hf1.NA))
```

Now, let's do it with `migration_matrix`.
So that migrants exit at the boundaries, we need to *not* normalize the matrix (`normalize=NULL`),
and compute the matrix for *all* cells, not just the ones that aren't missing 
(`from=1:length(habitat)`, or just construct the matrix from the habitat without missing values).
Note that if `normalize=NULL` then `migration_matrix` does not normalize the kernel to sum to a constant value,
but does multiply the entries by `area/sigma^2`, where area is the area of a cell.
In this case, this factor is equal to 1.
```{r simple_migmat}
M <- migration_matrix( habitat, kern=function(x) { ifelse(x>0,1/16,1/2) }, sigma=1, radius=1, normalize=NULL )
matrix(M[3,],nrow=nrow(habitat))
expect_equal( hmf, matrix(M%*%as.numeric(hmat),nrow=nrow(hmf)) )
expect_equal( hmf.NA, matrix(M%*%na_to_zero(as.numeric(hmat.NA)),nrow=nrow(hmf)) )
```

**Guts:** Absorbing external, not internal, boundaries
--------------------------------------------

Now, suppose we want migrants to exit at the external boundaries,
but not internal ones.
To make the internal boundaries reflecting,
in `focal()` we need to pad the matrix with zeros,
and use `fun=weighted.mean` instead of `fun=sum`, with `na.rm=TRUE`.
(This will be less efficient than with the default, `fun=sum`.)
Note that `focal` returns locally `fun(w*x)`, 
where `x` is the neighborhood of values,
so we need to replace `w` by the matrix of `1`s,
and use `weighted.mean` rather than `mean`.
```{r external_focal}
wm.fun <- function (x, na.rm) { weighted.mean(x,w=w,na.rm=na.rm) }
hf2 <- focal( habitat, w=(w>0), pad=TRUE, fun=wm.fun, na.rm=TRUE, padValue=0 )
hf2.NA <- focal( habitat.NA, w=(w>0), pad=TRUE, fun=wm.fun, na.rm=TRUE, padValue=0 )
layout(t(1:2))
plot(hf2,main="habitat",zlim=c(0,2))
plot(hf2.NA,main="habitat, with missings",zlim=c(0,2))
```

Now, let's do it with `migration_matrix`.
So that migrants do not exit at the internal boundaries, 
we need to normalize the matrix (`normalize=1`).
So that they *can* exit at the external boundaries,
we need to compute the migration matrix on a padded layer,
and then use the function `subset_migration` that restricts the
matrix to the part we want.
```{r external_migmat}
pad.extent <- extent(habitat)+c(-1,1,-1,1)*res(habitat) # extends by one cell
pad.habitat <- extend(habitat,pad.extent,value=0)
M <- migration_matrix( pad.habitat, kern=function(x) { ifelse(x>0,1/16,1/2) }, sigma=1, radius=1, normalize=1 )
M <- subset_migration( M, old=pad.habitat, new=habitat )
matrix(M[3,],nrow=nrow(habitat))

expect_equal( values(hf2), as.numeric(M%*%values(habitat)) )

pad.habitat.NA <- extend(habitat.NA,pad.extent,value=0)
M.NA <- migration_matrix( pad.habitat.NA, kern=function(x) { ifelse(x>0,1/16,1/2) }, sigma=1, radius=1, normalize=1 )
M.NA <- subset_migration( M.NA, old=pad.habitat.NA, new=habitat.NA  )

expect_equal( values(hf2.NA)[!is.na(values(habitat.NA))], as.numeric(M.NA%*%(values(habitat.NA)[!is.na(values(habitat.NA))])) )
```



**Guts:** Reflecting internal and external boundaries (non-boundaries)
-----------------------------------------------------------

Now, suppose we want migration to be conservative everywhere.
Now, in `focal()` we need to pad the matrix with `NA`s,
and use `fun=weighted.mean` instead of `fun=sum`, with `na.rm=TRUE`.
```{r internal_focal}
wm.fun <- function (x, na.rm) { weighted.mean(x,w=w,na.rm=na.rm) }
hf3 <- focal( habitat, w=(w>0), pad=TRUE, fun=wm.fun, na.rm=TRUE, padValue=NA )
hf3.NA <- focal( habitat.NA, w=(w>0), pad=TRUE, fun=wm.fun, na.rm=TRUE, padValue=NA )
layout(t(1:2))
plot(hf3,main="habitat",zlim=c(0,2))
plot(hf3.NA,main="habitat, with missings",zlim=c(0,2))
```

Doing this with `migration_matrix` is as easy as setting `normalize=1`.
```{r internal_migmat}
M <- migration_matrix( habitat, kern=function(x) { ifelse(x>0,1/16,1/2) }, sigma=1, radius=1, normalize=1 )
matrix(M[3,],nrow=nrow(habitat))
expect_equal( values(hf3), as.numeric(M%*%values(habitat)) )

M.NA <- migration_matrix( habitat.NA, kern=function(x) { ifelse(x>0,1/16,1/2) }, sigma=1, radius=1, normalize=1 )
expect_equal( values(hf3.NA)[!is.na(values(habitat.NA))], as.numeric(M.NA%*%(values(habitat.NA)[!is.na(values(habitat.NA))])) )
```



Less simple example
===================

Here's a simple habitat we'll work with.
Note that spatial units are in meters; the values are carrying capacities.
```{r setup_layer, fig.cap="The (random) habitat.", fig.width=fig.dim}
big.habitat <- raster(xmn=-1000, xmx=1000, ymn=-1000, ymx=1000, 
      resolution=100,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(big.habitat) <- sample( 100*c(1,2,5,NA), length(big.habitat), replace=TRUE )
plot(big.habitat)
```


Migration operations
====================

We can do migration in two ways:
either by matrix multiplication,
or by doing smoothing operations directly on the layers.


Matrices
--------

The easiest migration matrix to create is a nearest-neighbor migration:
```{r nn_mig}
nn.mig <- layer_adjacency(big.habitat)
image(nn.mig)
```

Alternatively, consider a Gaussian kernel:
```{r gaus_mig}
gauss.mig <- migration_matrix(big.habitat,sigma=100,radius=500,kern="gaussian")
image(gauss.mig)
```

Or, a two-dimensional, symmetric Cauchy kernel:
```{r cauchy_mig}
cauchy.mig <- migration_matrix(big.habitat,sigma=100,radius=800,kern="cauchy")
image(cauchy.mig)
```


Smoothing operations
--------------------

The `raster` package does smoothing operations directly, with `raster::focal`.
For instance, we can assign random values to the non-NA values in the layer
and then smooth them with a Gaussian kernel:
```{r gauss_smooth}
N <- big.habitat
values(N)[!is.na(values(N))] <- rpois(sum(!is.na(values(N))),values(big.habitat)[!is.na(values(N))])
plot(N)
gauss.N <- migration_matrix(N,radius=500,sigma=100,kern="gaussian")
plot(gauss.N)
```


Discretization
==============

The naive method for constructing the migration matrix
from a migration kernel $k$ and scale $\sigma$
has the migration rate from a cell centered at $x$ to a cell centered at $y$
equal to $k(\frac{y-x}{\sigma})$.
However, if $\sigma$ is of the same size or smaller than $|x-y|$, this is not a good approximation
to the corresponding continuous model.
The migration rate between squares $S_x$ and $S_y$
is actually proportional to
$$  \int_{S_x} \int_{S_y} k\left(\frac{u-v}{\sigma}\right) du dv .  $$

Note that even if the initial kernel only depends on distance,
the resulting discretized one may not.
Let's check what this looks like,
summing up migration rates from the cell at zero to other cells
for a Gaussian kernel with a $\sigma$ equal to one-half the cell size.
```{r show_discretized}
naive.migration <- migration( kern="gaussian",
                             sigma=0.5,
                             radius=5,
                             normalize=NULL )
center.loc <- as.integer(cellFromXY( habitat, c(0,0) ))
habitat.fine <- disaggregate(habitat, fact=10)
fine.locs <- xyFromCell(habitat.fine,seq_len(prod(dim(habitat.fine))))
fine.in.coarse <- cellFromXY(habitat,fine.locs)
M.fine <- migration_matrix( habitat.fine, 
                           kern=naive.migration$kern, 
                           sigma=naive.migration$sigma, 
                           radius=naive.migration$radius, 
                           normalize=naive.migration$normalize,
                           from=which(fine.in.coarse==center.loc), to=seq_len(prod(dim(habitat.fine))) )
M.aggr <- aggregate_migration( M.fine, old=habitat.fine, new=habitat,
                           from.old=which(fine.in.coarse==center.loc), to.old=seq_len(prod(dim(habitat.fine))),
                           from.new=center.loc, to.new=seq_len(prod(dim(habitat))) )
M.coarse <- migration_matrix( habitat,
                            kern=naive.migration$kern, 
                            sigma=naive.migration$sigma, 
                            radius=naive.migration$radius, 
                            normalize=naive.migration$normalize,
                            from=center.loc, to=seq_len(prod(dim(habitat))) )
plm <- function (x,...) { z <- habitat; values(z) <- x; plot(z,...) }
# the difference in the matrices
layout(t(1:2))
plm(M.aggr[1,],main="aggregated",zlim=range(M.aggr@x,M.coarse@x))
plm(M.coarse[1,],main="coarse",zlim=range(M.aggr@x,M.coarse@x))
plm((M.coarse-M.aggr)[1,],main="error")
plm((abs(M.coarse-M.aggr)/M.aggr)[1,],main="relative error")
# now, as a function of distance
distvec <- values( distanceFromPoints( habitat, xyFromCell(habitat,cell=center.loc) ) )
matplot( distvec, cbind( M.aggr[1,], M.coarse[1,] ) )
matplot( distvec, cbind( M.aggr[1,], M.coarse[1,] ), log='y' )
```

Our strategy for producing an approximately correct migration matrix
without actually integrating over all pairs of cells
will be to construct an approximating function of distance as with the procedure above on the migration stencil
and to use this as before.
This is implemented with the function `discretize kernel`:
```{r disc_kernel}
diskern <- discretize_kernel( get_kernel(naive.migration$kern), res=res(habitat), 
                             radius=naive.migration$radius, sigma=naive.migration$sigma )
M.dk <- migration_matrix( habitat, 
                        kern=diskern,
                        sigma=naive.migration$sigma, 
                        radius=naive.migration$radius, 
                        normalize=naive.migration$normalize,
                        from=center.loc, to=seq_len(prod(dim(habitat))) )
matplot( distvec, cbind( M.aggr[1,], M.coarse[1,], M.dk[1,] ), xlab='distance', ylab='migration probability' )
xx <- seq(0,max(distvec),length.out=400)
yy <- diskern(xx/naive.migration$sigma) * prod(attr(diskern,"res")/naive.migration$sigma)
lines( xx, yy, col=3 )
legend("topright",col=1:3,lty=1:3,legend=c("aggregated","coarse","discretized coarse"))
matplot( distvec, cbind( M.aggr[1,], M.coarse[1,], M.dk[1,] ), log='y', xlab='distance', ylab='migration probability' )
lines( xx, yy, col=3 )
```


