---
title: "Migration methos"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


Here's a simple habitat we'll work with.
Note that spatial units are in meters; the values are carrying capacities.
```{r setup_layer, fig.cap="The (random) habitat.", fig.width=fig.dim}
habitat <- raster(xmn=-1000, xmx=1000, ymn=-1000, ymx=1000, 
      resolution=100,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(habitat) <- sample( 100*c(1,2,5,NA), length(habitat), replace=TRUE )
plot(habitat)
```


Migration operations
====================

We can do migration in two ways:
either by matrix multiplication,
or by doing smoothing operations directly on the layers.


Matrices
--------

The easiest migration matrix to create is a nearest-neighbor migration:
```{r nn_mig}
nn.mig <- layer_adjacency(habitat)
image(nn.mig)
```

Alternatively, consider a Gaussian kernel:
```{r gaus_mig}
gauss.mig <- kernel_adjacency(habitat,sigma=100,radius=500,fun="gaussian")
image(gauss.mig)
```

Or, a two-dimensional, symmetric Cauchy kernel:
```{r cauchy_mig}
cauchy.mig <- kernel_adjacency(habitat,sigma=100,radius=800,fun="cauchy")
image(cauchy.mig)
```


Smoothing operations
--------------------

The `raster` package does smoothing operations directly, with `raster::focal`.
For instance, we can assign random values to the non-NA values in the layer
and then smooth them with a Gaussian kernel:
```{r gauss_smooth}
N <- habitat
values(N)[!is.na(values(N))] <- rpois(sum(!is.na(values(N))),values(habitat)[!is.na(values(N))])
plot(N)
gauss.N <- kernel_migrate(N,radius=500,sigma=100)
plot(gauss.N)
```
