---
title: "Package overview"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(jsonlite)
devtools::load_all("..")
```


To run a simulation, we need to know
where the species lives,
how it disperses,
and how it reproduces and dies.
Given a demographic model of this sort,
we then keep track of the numbers of various genotypes
across the species range.

These are divided among different objects as follows:


`population`
----------

Class **population**:

`habitat`

: a Raster with values that can be used in other computations.

`accessible`

: vector of indices of cells in `habitat` that migrants will attempt to go to

`habitable`

: vector of indices of cells in `habitat` that may have positive population

`genotypes`

: character vector of the genotypes

`N`

: matrix indexed by (habitable cells) x (genotypes) giving the number of each genotype in each habitable cell


The configuration can be stored as JSON:

`habitat`

: The path to the file where the raster is stored.

`inaccessible.value`

: The values in the raster that should be marked as inaccessible.

`uninhabitable.value`

: The values in the raster that should be marked as not habitable.

`genotypes`

: A character vector of genotypes.


**Example:**

Here is the JSON:
```{r pop_json, echo=FALSE}
pop.list <- list(
                 habitat = system.file("inst/extdata/test_raster.gri",package="landsim"),
                 inaccessible.value = "NA",
                 uninhabitable.value = 0.0,
                 genotypes = c( "aa", "aA", "AA" )
             )
pop.json <- toJSON(pop.list,prett=TRUE)
pop.json
```

This can be parsed using `make_population()`:
```
pop <- make_population(pop.json)
```


`migration`
---------

One structure we use several times
is the setup for smoothing (with `migrate`);
for this we need to know:

Class **migration**:

`kern`

: function that gives weights for neighboring cells as a function of Euclidean distance

`sigma`

: scaling factor on distance before being passed to `kern`

`radius`

: maximum range of the smoother

`normalize`

: normalization factor applied to the smoother


**Example:**

Here is the JSON:
```{r pop_json, echo=FALSE}
migr.list <- list(
                 kern = "gaussian",
                 sigma = 100,
                 radius = 1000,
                 normalize = 1.0
             )
migr.json <- toJSON(migr.list,prett=TRUE)
migr.json
```

This can be parsed using `make_population()`:
```
migr <- make_migration(migr.json)
```


migration.matrix
----------------

A `migration.matrix` object extends a `migration` object,
in that it has a precomputed migration matrix,
and hence is tied to a particular `population` setup
(although this is not included).  This has additional entries

Class **migration.matrix**:

`M`

: The pre-computed migration matrix.


**Example:**

This is added with the function `migration`:

```
migr <- migration( migr, population=pop, do.M=TRUE )
```


`vital_rate`
----------

Vital rates may be numbers or vectors
but may also be functions,
to allow them to depend on the current state of the system.

**Example:**

Here is the JSON:
```{r vital_json, echo=FALSE}
vital.list <- list(
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             carrying.capacity = 10*habitat,
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1,
                                     do.M=TRUE,
                                     population=population
                                 ),
             value = function (N, ...) {
                out <- r0 / ( 1 + migrate(competition,x=rowSums(N))/carrying.capacity )
                return( cbind( aa=out, aA=s*out, AA=s^2*out ) )
             }
         )

vital.json <- '
{
    "value" : { "function" : 
              " out <- r0 / ( 1 + migrate(competition,x=rowSums(N))/carrying.capacity ) ;
                return( cbind( aa=out, aA=s*out, AA=s^2*out ) ) " 
        },
    "r0" : 0.01,
    "s" : 1.5,
    "carrying.capacity" : { "R" : "10*habitat" },
    "competition" : { "migration" : { 
            "kern" : "gaussian",
            "sigma" : 100,
            "radius" : 300,
            "normalize" : 1.0
        } }
}
'
```



`demography`
----------

Several of the base demographic parameters we might want to specify
(as a fixed number or a layer),
or compute (as a function of the current population state).
Such **parameters** can be either something numeric
or else a function that is applied to `N` to get the result.

To do a single generation we need to know:

Class **demography**:

`prob.seed`

: `parameter` object for  probability of seeding per individual per year

`fecundity`

: `parameter` object for mean number of seeds per seeding individual per year

`prob.germination`

: `parameter` object for probability of germination

`prob.survival`

: `parameter` object for probability of survival of already existing individuals

`pollen.migration`

: `migration` object for pollen dispersal (normalized to total pollen production)

`seed.migration`

: `migration` object for seed dispersal (normalized to 1)

`genotypes`

: vector of names of the genotypes represented by the layers

`mating`

: genotype x genotype x genotype tensor, with entry
    `[i,j,k]` the probability that genotypes `i` and `j` combine to make offspring genotype `k`




**Example:**

Here is the JSON:
```{r demog, echo=FALSE}
demog.list <- list(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = gsub("\n","",vital.json),
        prob.survival = 0.9, 
        pollen.migration = migr.list, 
        seed.migration = migr.list, 
        genotypes = c("aa", "aA", "AA")
     )
demog.json <- toJSON(demog.list,pretty=TRUE)
demog.json
```
