---
title: "Package overview"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r doc_setup, include=FALSE}
fig.dim <- 4
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(landsim)
```
```{r test}
f <- vital( function (N,...) { N+x }, x=3 )
f(10)
f$x
f$x <- 12
f(10)
```


To run a simulation, we need to know
where the species lives,
how it disperses,
and how it reproduces and dies.
Given a demographic model of this sort,
we then keep track of the numbers of various genotypes
across the species range.

These are divided among different objects as follows:


`population`
----------

Class **population**:

`habitat`

: a Raster with values that can be used in other computations.

`accessible`

: vector of indices of cells in `habitat` that migrants will attempt to go to

`habitable`

: vector of indices of cells in `habitat` that may have positive population

`genotypes`

: character vector of the genotypes

`N`

: matrix indexed by (habitable cells) x (genotypes) giving the number of each genotype in each habitable cell


**Set-up:**

This can be initiated with `make_population()` by specifying the following:

`habitat`

: The path to the file where the raster is stored.

`inaccessible.value`

: The values in the raster that should be marked as inaccessible.

`uninhabitable.value`

: The values in the raster that should be marked as not habitable.

`genotypes`

: A character vector of genotypes.


**Example:**

```{r pop_defn}
pop <- make_population(
                 habitat = system.file("extdata/test_raster.gri",package="landsim"),
                 inaccessible.value = "NA",
                 uninhabitable.value = 0.0,
                 genotypes = c( "aa", "aA", "AA" )
             )
```


`migration`
---------

One structure we use several times
is the setup for smoothing (with `migrate`);
for this we need to know:

Class **migration**:

`kern`

: function that gives weights for neighboring cells as a function of Euclidean distance

`sigma`

: scaling factor on distance before being passed to `kern`

`radius`

: maximum range of the smoother

`normalize`

: normalization factor applied to the smoother

`n`

: number of times to apply the smoother to produce one "migration" step

**Example:**

```{r migr_defn}
migr <- migration(
                 kern = "gaussian",
                 sigma = 100,
                 radius = 1000,
                 normalize = 1.0
             )
```


migration.matrix
----------------

A `migration.matrix` object extends a `migration` object,
in that it has a precomputed migration matrix,
and hence is tied to a particular `population` setup
(although this is not included).  This has additional entries

Class **migration.matrix**:

`M`

: The pre-computed migration matrix.


**Example:**

This is added with the function `setup_migration()`.

```
migr <- setup_migration( migr, population=pop )
```


`vital_rate`
----------

Vital rates may be numbers or vectors
but may also be functions,
to allow them to depend on the current state of the system.
We would like to have the parameters in these visible,
changeable, and carried around with them.
To this end, we define `vital` objects
by specifying a function
and auxialliary parameters that will be kept along with the function.

**Simple example:**

This class allows assigning and accessing things directly from the environment of the function,
without having them clutter up the global environment.
```{r simple_vital_defn}
f <- vital( function (N,...) { N+x }, x=3 )
f(10)
f$x
f$x <- 12
f(10)
```



**Example:**

```{r vital_defn}
germ.vital <- vital(
             function (N, ...) {
                out <- r0 / ( 1 + migrate(rowSums(N),competition)/carrying.capacity )
                return( cbind( aa=out, aA=s*out, AA=s^2*out ) )
             },
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             carrying.capacity = 10,
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1
                                 )
         )
```

Note that this also has a `migration` object as part of its definition.
This needs to be set up also:
```{r setup_vital}
germ.vital <- setup_vital(germ.vital,pop)
```


`demography`
----------

Several of the base demographic parameters we might want to specify
(as a fixed number or a layer),
or compute (as a function of the current population state).
Such **parameters** can be either something numeric
or else a function that is applied to `N` to get the result.

To do a single generation we need to know:

Class **demography**:

`prob.seed`

: `parameter` object for  probability of seeding per individual per year

`fecundity`

: `parameter` object for mean number of seeds per seeding individual per year

`prob.germination`

: `parameter` object for probability of germination

`prob.survival`

: `parameter` object for probability of survival of already existing individuals

`pollen.migration`

: `migration` object for pollen dispersal (normalized to total pollen production)

`seed.migration`

: `migration` object for seed dispersal (normalized to 1)

`genotypes`

: vector of names of the genotypes represented by the layers

`mating`

: genotype x genotype x genotype tensor, with entry
    `[i,j,k]` the probability that genotypes `i` and `j` combine to make offspring genotype `k`




**Example:**

```{r demog}
demog <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germ.vital,
        prob.survival = 0.9, 
        seed.migration = migr,
        pollen.migration = migration(
                         kern="gaussian",
                         sigma=300,
                         radius=900,
                         normalize=10
                     ),
        genotypes = c("aa", "aA", "AA")
     )
```

This can be set up for use with a population with `setup_demography`,
which sets up all the associated migration and vital objects.
```
demog <- setup_demography(demog,pop)
```
