---
title: "Simulations on large rasters."
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r doc_setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(microbenchmark)
devtools::load_all("..")
set.seed(42)
this.plot <- function (x,nr=1,zlim=c(0,max(values(x),na.rm=TRUE)),...) {
    plot(x,nr=nr,zlim=zlim,...)
}
```

**How well do the simulation methods work with larger rasters?**

Here are the demographic parameters:
```{r fancy_setup}
germination.args <- list( 
             r0 = 0.01,
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1
                                 )
         )
germination_fun <- function (N,carrying.capacity,...) {
    germination.args$r0 / ( 1 + migrate_raster(N,germination.args$competition)/carrying.capacity )
}

this.demography <- demography(
        prob.seed = 0.2,
        fecundity = 200,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 100,
                            radius = 1000,
                            normalize = NULL
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 20,
                            radius = 400,
                            normalize = 1
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```

Raster extent
=============

We will time a generation on rasters of different sizes.
Here is a function to create a $k$-kilometer square random raster.
(But note that the base units of the raster are in meters.)
```{r habitat_setup}
habitat_size <- function (k) {
    habitat <- raster(xmn=-k*1000/2, xmx=k*1000/2, ymn=-k*1000/2, ymx=k*1000/2, 
          resolution=100, #fixed resolution
          crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
    values(habitat) <- sample( 100*c(1,2,5,NA), length(habitat), replace=TRUE )
    migrate_raster(habitat,kern="gaussian",sigma=200,radius=1000)
}
```

Here's timing for one generation on rasters of various sizes:
```{r init_new_pop}
sizes <- 2^(0:6)
size.results <- lapply( sizes, function (k) {
                  habitat <- habitat_size(k)
                  N <- rpois_raster( do.call( brick, list(habitat)[rep(1,length(this.demography$genotypes))] ) )
                  names(N) <- this.demography$genotypes
                  gen.time <- system.time( NN <- generation_raster(N,this.demography,carrying.capacity=habitat) )
                  return( list( time=gen.time, N=N, NN=NN ) )
          } )
```

Here's the results:
```{r show_timing_results, fig.width=2*fig.dim}
size.tab <- data.frame( area=sizes^2, ncell=sapply(lapply(size.results,"[[","N"),ncell), t(sapply( size.results, "[[", "time" )) )
size.tab
layout(t(1:2))
with(size.tab, {
     plot(area, user.self, log='xy', xlab='total area', ylab='computation time (sec)') 
     plot(ncell, user.self, log='xy', xlab='number of raster cells', ylab='computation time (sec)') 
     } )
```

The computation time will depend critically on the number of cells within the radii
of the `migration` operations;
but at these settings, adding
`r round( 1/coef(with( subset(size.tab,ncell>5e3), lm(user.self ~ ncell) ))[2], 0 )`
cells increases the run time by a second.


Raster resolution
=================

Changing the resolution at a fixed size also changes the number of cells,
but affects the running time quite differently.
Here is a function to create a 10-kilometer square random raster, with different resolutions.
(But note that the base units of the raster are in meters.)
```{r habitat_setup_res}
habitat_res <- function (resolution) {
    habitat <- raster(xmn=-10*1000/2, xmx=10*1000/2, ymn=-10*1000/2, ymx=10*1000/2, 
          resolution=resolution,
          crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
    values(habitat) <- sample( 100*c(1,2,5,NA), length(habitat), replace=TRUE )
    migrate_raster(habitat,kern="gaussian",sigma=200,radius=1000)
}
```

Here's timing for one generation on rasters of various sizes:
```{r run_res}
resolutions <- 100*2^(-2:2)
res.results <- lapply( resolutions, function (resolution) {
                  habitat <- habitat_res(resolution)
                  N <- rpois_raster( do.call( brick, list(habitat)[rep(1,length(this.demography$genotypes))] ) )
                  names(N) <- this.demography$genotypes
                  gen.time <- system.time( NN <- generation_raster(N,this.demography,carrying.capacity=habitat) )
                  return( list( time=gen.time, N=N, NN=NN ) )
          } )
```

Here's the results:
```{r show_res_results, fig.width=2*fig.dim}
res.tab <- data.frame( resolution=resolutions, ncell=sapply(lapply(res.results,"[[","N"),ncell), t(sapply( res.results, "[[", "time" )) )
res.tab
layout(t(1:2))
with(res.tab, {
     plot(resolution, user.self, log='xy', xlab='total area', ylab='computation time (sec)') 
     plot(ncell, user.self, log='xy', xlab='number of raster cells', ylab='computation time (sec)') 
     } )
```

The computation time will depend critically on the number of cells within the radii
of the `migration` operations;
but at these settings, adding
`r round( 1/coef(with( subset(res.tab,ncell>5e3), lm(user.self ~ ncell) ))[2], 0 )`
cells increases the run time by a second.


Two implementations
===================

Based on the above, we have two implementations of the method,
one that works with `Raster` objects (so can take advantage of methods for those that work on layers too big to fit in memory);
and one that precomputes a migration matrix and works directly with numeric matrices.

Let's compare the speeds.
```{r compare_methods}
habitat <- habitat_size(6)
N <- rpois_raster( do.call( brick, list(habitat)[rep(1,length(this.demography$genotypes))] ) )
names(N) <- this.demography$genotypes
raster.time <- microbenchmark( generation_raster(N,this.demography,carrying.capacity=habitat), times=10 )

pop <- population( 
                  habitat = habitat,
                  genotypes = this.demography$genotypes,
                  N = matrix(values(N)[!is.na(values(habitat))],ncol=3)
             )
matrix.demography <- this.demography
matrix.demography$seed.migration <- migration( matrix.demography$seed.migration, do.M=TRUE, layer=habitat )
matrix.demography$pollen.migration <- migration( matrix.demography$pollen.migration, do.M=TRUE, layer=habitat )
matrix.germination.args <- germination.args
matrix.germination.args$competition <- migration( matrix.germination.args$competition, do.M=TRUE, layer=habitat )
matrix.demography$prob.germination <- function (N,carrying.capacity,...) {
    matrix.germination.args$r0 / ( 1 + migrate(matrix.germination.args$competition,N)/carrying.capacity )
}
matrix.time <- microbenchmark( generation(pop,matrix.demography,carrying.capacity=values(habitat)[!is.na(values(habitat))]), times=10 )

rbind( raster.time, matrix.time )
```

That's a speedup of `r round(mean(raster.time$time)/mean(matrix.time$time))` times, pretty good.
