---
title: "Methods for sampling lineages"
author: "Peter Ralph"
date: "`r Sys.Date()`"
---


```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(landsim)
set.seed(42)
```

# Set-up

Here is the habitat we will work with.
Note that units are in meters, and the resolution of the raster is 100m.
```{r setup_layer, fig.cap="The (random) habitat.", fig.height=4, fig.width=4}
pop <- make_population(
            habitat = random_habitat(),
            inaccessible.value = NA,
            uninhabitable.value = NA,
            genotypes = c("aa","aA","AA"),
            N = 0
        )
pop$N[,"aa"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/4)
pop$N[,"aA"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/2)
pop$N[,"AA"] <- rpois(nrow(pop$N),values(pop$habitat)[pop$habitable]/4)
plot(pop$habitat)
```

Here is the basic, default demography,
modified to have population regulation through germination probabilities:
```{r basic_demog}
basic.migr <- migration(
                    kern = "gaussian",
                    sigma = 300,
                    radius = 1000,
                    normalize = 1
             )
basic.demog <- demography(
        prob.seed = 0.05,
        fecundity = 200,
        prob.germination = 0.4,
        prob.survival = 0.6,
        pollen.migration = basic.migr,
        seed.migration = basic.migr,
        genotypes = c("aa","aA","AA")
    )

demog <- basic.demog
demog$prob.germination <- vital(
                    function (N,...) {
                        out <- r0 / ( 1 + migrate(competition,x=rowSums(N))/K )
                        cbind( aa=out, aA=out, AA=out )
                    },
                    r0 = 0.4,
                    K = values(pop$habitat)[pop$habitable]/5,
                    competition = migration(
                                kern="gaussian",
                                sigma=200,
                                radius=400,
                                normalize=1
                        )
                )
```


Here are the total numbers of the three genotypes:
```{r plot_prob_germ}
demog <- setup_demography( demog, pop )
sim <- simulate_pop( pop, demog, times=seq(0,100,length.out=101),
                summaries=list( totals=function(N){colSums(N)} )
            )
matplot(sim$summaries[["totals"]],type='l',lty=1, log='y', ylab="number of individuals")
legend("bottomright",lty=1,col=1:3,legend=pop$genotypes)
```

# Lineages

Now we have a population history and the demographic model that produced it.
Let's start a bit earlier.
We can ask the `generation` function to return *everything*,
not just total numbers, namely:

1. `seeders` : number of individuals in each location that produce seeds
2. `pollen` : *expected* pollen flux (this is `pollen.migration` applied to `N`; since this is in expected numbers we can treat pollen as diploid, averaging over meiosis later)
3. `seed.production` : *expected* numbers of seeds of each genotype produced by mating between pollen and local genotypes (using the `mating` tensor)
4. `seeds.dispersed` : `seed.production` with `seed.migration` applied to it (still *expected* numbers)
5. `germination` : numbers of new individuals germinating in each location
6. `death` : number of previously present individuals that have died

The numbers of genotypes (`N`) in the next generation are `survivors + germination`,
where `survivors = N - death`.

Lineages each are labeled by an *allele*, not a genotype;
to keep things simple, we will follow lineages of a single allele; for the discussion below this will be allele $A$.
Suppose there genotype proportions among the seed parents are $p_{AA}$, $p_{Aa}$, and $p_{aa}$;
they are $q_{AA}$, $q_{Aa}$, and $q_{aa}$ among the pollen parents,
the probability that parents $u$ and $v$ produce offspring genotype $g$ is $t_{u,v,g}$.
Lineages in $AA$ offspring choose pollen or seed parent with equal probability (1/2 each).
The numbers of $Aa$ offspring coming from (seed,pollen) parents $(u,v)$ is proportional to $p_u q_v t(u,v,Aa)$;
an $A$ lineage in an $Aa$ offspring will choose a pair of parents $(u,v)$ with probability proportional to this.
Let $A(u)$ be the number of $A$ alleles in genotype $u$.
Assuming that meiosis is fair, given parents $(u,v)$ the $A$ lineage will choose parent $u$ with probability $A(u)/(A(u)+A(v))$.

Therefore,
to sample lineage movement backwards through this we just need to:

0.  Choose genotypes for the lineages to be in.
1.  Stay in the same place with probability `survivors/(survivors+germination)`.
2.  Otherwise, do a backwards migration step with `seed.migration` weighted by `seed.production`, then
3.  choose whether to follow a seed or pollen parent, and their genotype, by a backwards step through the `mating` tensor as described above,
    using `seeders` and `pollen` as weights.
4.  If the seed parent is chosen, stay put.
5.  If the pollen parent is chosen, do another backwards migration step with `pollen.migration` weighted by `N`.

These steps are implemented in `lineage_generation`;
here we apply them
starting with a sample of lineages chosen uniformly from the possible $A$ gametes to make sure that things make sense.

```{r one_lineage_step, fig.width=fig.dim}

one.gen <- generation( pop, demog, return.everything=TRUE )

# current generation
now.N <- pop$N - one.gen$death + one.gen$germination

# choose lineage locations: coordinates are (genotype, location index)
mean.n.lineages <- 1000
n.lineages <- rpois_matrix( sweep(now.N,2,c(0,1,2),"*")*mean.n.lineages/sum(now.N) )
lineages <- arrayInd( rep( seq_along(n.lineages), n.lineages ), .dim=dim(now.N) )
colnames(lineages) <- c("location","genotype")

parents <- lineage_generation( lineages, N=pop$N, gen=one.gen, num.alleles=c(0,1,2) )

now.locs <- xyFromCell(pop$habitat,which(pop$habitable)[lineages[,1]])
parent.locs <- xyFromCell(pop$habitat,which(pop$habitable)[parents[,1]])

plot(pop$habitat)
suppressWarnings( arrows( x0=now.locs[,'x'], x1=parent.locs[,'x'], 
           y0=now.locs[,'y'], y1=parent.locs[,'y'], 
           length=0.1 ) )

```

Now to do this across many generations we need to keep track of everything across those generations.
Here are a few lineages for 100 generations:
```{r many_lineage_steps, fig.width=fig.dim}

many.gens <- simulate_pop( pop, demog, times=1:100, tinit=1, return.everything=TRUE )
mean.n.lineages <- 20
now.N <- many.gens$N[,,length(many.gens$times)]
n.lineages <- rpois_matrix( sweep(now.N,2,c(0,1,2),"*")*mean.n.lineages/sum(now.N) )
lineages <- arrayInd( rep( seq_along(n.lineages), n.lineages ), .dim=dim(now.N) )
colnames(lineages) <- c("location","genotype")

lin.hist <- simulate_lineages( lineages, many.gens, num.alleles=c(0,1,2) )

lin.locs <- xyFromCell(pop$habitat,which(pop$habitable)[lin.hist[,1,]])
dim(lin.locs) <- c( dim(lin.hist)[c(1,3)], 2 )

plot(pop$habitat)
for (k in 1:nrow(lin.locs)) {
    now.locs <- lin.locs[k,-1,]
    parent.locs <- lin.locs[k,-dim(lin.locs)[2],]
    do.arrows <- ( rowSums( abs(now.locs-parent.locs) )>0 )
    suppressWarnings( arrows( x0=now.locs[do.arrows,1], x1=parent.locs[do.arrows,1], 
               y0=now.locs[do.arrows,2], y1=parent.locs[do.arrows,2], 
               length=0.1, col=rainbow(nrow(lin.locs))[k] ) )
}

```

Note that these lineages are all of type A, but are jumping back and forth between Aa and AA genotypes;
we're not looking at that.
