---
title: "Simulating a simple selective sweep on a landscape."
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulating selective sweeps on a landscape}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r doc_setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
library(jsonlite)
devtools::load_all("..")
set.seed(42)
```


Here's a simple layer we'll work with.
Note that units are in meters.
```{r setup_layer, fig.cap="The (random) habitat.", fig.height=4, fig.width=4}
habitat <- raster(xmn=-1000, xmx=1000, ymn=-1000, ymx=1000, 
      resolution=100,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(habitat) <- exp(2+rnorm(ncell(habitat)))
habitat <- migrate_raster( habitat, kern="gaussian", sigma=100, radius=1500 )
plot(habitat)
```

The total carrying capacity is `r sum(values(habitat),na.rm=TRUE)`.

```{r plot_setup}
# plotting function for typical layers below
pl <- function (x,v,nr=1,zlim, ...) {
    if (!missing(v)) {
        if (nlayers(x)<NCOL(v)) {
            x <- do.call(stack,list(x)[rep(1,NCOL(v))])
        }
        values(x)[!is.na(values(x))] <- as.numeric(v)
    }
    if (missing(zlim)) {
        if (inherits(x,"Raster")) {
            zlim  <- range(0,values(x),finite=TRUE)
        } else if (inherits(x,"population")) {
            zlim  <- range(0,as.numeric(x$N),finite=TRUE)
        }
    }
    plot(x,nr=nr,zlim=zlim,...)
}
```

Genotype-dependent recruitment
------------------------------

Let's suppose there's strong, additive selection:
heterozygous *Aa* alleles are 10% more likely to germinate than *aa* alleles,
and *AA* alleles are 20% more likely to germinate.
The `germination_fun` gets called with the current state of the population as an argument (`N`),
whose result is multiplied by the total seed flux, per genotype and per location,
to get the mean number of newly recruited individuals.


Here's the demographic set-up
```{r demog_setup}
germination.args <- list( 
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1,
                                     do.M=TRUE,
                                     population=habitat
                                 )
         )
germination_fun <- function (N, carrying.capacity, ...) {
    out <- germination.args$r0 / ( 1 + migrate(germination.args$competition,x=rowSums(N))/carrying.capacity )
    return( cbind( aa=out, aA=germination.args$s*out, AA=germination.args$s^2*out ) )
}

this.demography <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 100,
                            radius = 1000,
                            normalize = NULL,
                            do.M=TRUE,
                            population=habitat
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 20,
                            radius = 400,
                            normalize = 1,
                            do.M=TRUE,
                            population=habitat
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```



We'll start the population with only a few of the advantageous alleles:
```{r init_pop}
total.habitat <- sum(values(habitat),na.rm=TRUE)
pop <- population( 
                  habitat = habitat,
                  genotypes = c("aa","aA","AA"),
                  N = cbind( aa=rpois_raster(habitat,only.values=TRUE),
                             aA=rpois_raster(habitat*100/total.habitat,only.values=TRUE),
                             AA=0 )
             )

pl(pop)
```

Let's check this looks to allow a stable population.
The mean number of offspring per individual
the probability of seeding
multiplied by the fecundity,
multiplied by the probability of germination;
this must be greater than the probability of death,
calculated at low density.
Here this is:
```{r stability_check}
exp.pop <- pop
exp.pop$N[] <- 1
vitals <- generation(exp.pop,this.demography,
                     carrying.capacity=values(pop$habitat)[pop$habitable],
                     expected=TRUE,return.everything=TRUE)
for (k in seq_along(vitals)) {
    pl(exp.pop$habitat,v=vitals[[k]],main=c("",paste(names(vitals)[k],"density 1"),""))
}
```

Just for comparison, here's the number of germinating seeds per capita at a much higher density
```{r high_density}
exp.pop$N[] <- 500
vitals <- generation(exp.pop,this.demography,
                     carrying.capacity=values(pop$habitat)[pop$habitable],
                     expected=TRUE,return.everything=TRUE)
pl(exp.pop$habitat,v=vitals$germination/exp.pop$N,main=c("","per capita germination at density 500","") )
```

Now, we can run the simulation.
Note that we have to pass `carrying.capacity` in.
We'll record the full state of the population at several times, 
and record the total number of each genotype in every generation.
```{r run_sim}
plot.times <- 1+c(1,10,20,40,100,200)
sim <- simulate( pop, this.demography, times=plot.times, 
                carrying.capacity=values(pop$habitat)[pop$habitable],
                summaries=list( totals=function(x){colSums(x$N)} )
            )
```

Here is a video of the simulation:
```{r plot_sim_snapshots, fig.show="animate", aniopts="controls", interval=0.2}
for (k in seq_along(plot.times)) {
    pl(pop$habitat,v=sim$N[,,k],main=c("",sprintf("t=%d",plot.times[k])),zlim=range(sim$N,finite=TRUE))
}
```

And, here are the total numbers of the three genotypes:
```{r plot_census,fig.height=1.5*fig.dim}
matplot( sim$summaries[["totals"]], type='l', xlab='time', ylab='numbers', lty=1 )
legend("topright",lty=1,col=1:3,legend=colnames(sim$summaries[["totals"]]))
```
