---
title: "Sierpinski Serpents"
author: "Peter Ralph"
date: "`r Sys.Date()`"
output: 
    rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Sierpinski Serpents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



```{r doc_setup, include=FALSE}
fig.dim <- 3
knitr::opts_chunk$set(fig.width=3*fig.dim,fig.height=fig.dim,fig.align='center')
library(Matrix)
library(raster)
devtools::load_all("..")
set.seed(42)
```

Suppose, just for fun, that we have a population living on the [Sierpinski gasket](http://en.wikipedia.org/Sierpinski_gasket).
There's a handy function to create this:
```{r setup_habitat}
diam <- 1e4
habitat <- raster(xmn=-diam, xmx=diam, ymn=-diam, ymx=diam, 
      resolution=100,
      crs="+proj=utm +zone=11 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")
values(habitat) <- 1.0
sierp <- sierpinski_overlay(habitat,random=TRUE,n=3)
plot(sierp)
```

Demography
==========

Here's the demographic set-up:
```{r demog_setup}
germination.args <- list( 
             r0 = 0.01,  # one in ten seeds will germinate at low densities
             s = 1.5,    # multiplicative selective benefit of the A allele
             competition = migration(
                                     kern="gaussian",
                                     sigma=100,
                                     radius=300,
                                     normalize=1,
                                     do.M=TRUE,
                                     population=habitat
                                 )
         )
germination_fun <- function (N, carrying.capacity, ...) {
    out <- germination.args$r0 / ( 1 + migrate(germination.args$competition,x=rowSums(N))/carrying.capacity )
    return( cbind( aa=out, aA=germination.args$s*out, AA=germination.args$s^2*out ) )
}

this.demography <- demography(
        prob.seed = 0.2,
        fecundity = 100,
        prob.germination = germination_fun,
        prob.survival = 0.9,
        pollen.migration = migration(
                            kern = function (x) { exp(-sqrt(x)) },
                            sigma = 30,
                            radius = 500,
                            normalize = NULL,
                            do.M=TRUE,
                            population=habitat
                     ),
        seed.migration = migration(
                            kern = "gaussian",
                            sigma = 50,
                            radius = 400,
                            normalize = 1,
                            do.M=TRUE,
                            population=habitat
                     ),
        genotypes = c("aa","aA","AA"),
        mating = mating_tensor( c("aa","aA","AA") )
    )
```




Probability of establishment
============================


The probability of establishment only depends on what happens relatively nearby.
We'll select random neighborhoods,
seed them with mutants,
and see how many establish.


Wave speed
==========


